\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage[margin=0.7in]{geometry}

\begin{document}

\title{Formal methods at work 2018}
\author{Renan Greca; Konstantin Prokopchik}

\maketitle

\begin{enumerate}
	\item Consider two statements, $S_1$ and $S_2$, that are equivalent in big step semantics ($S_1 \approx_{BS} S_2$).
		For every program $S$ and memory state $
	\sigma$, it is also true that $(S_1;S) \approx_{BS} (S_2;S)$.
		Consider the following transition: $(S_1;S,\sigma)\Rightarrow\sigma'$: by the definition of the big-step semantics for sequential operations, there is a memory state $\sigma''$ such that $(S_1,\sigma)\Rightarrow\sigma''$:
		
		\begin{equation*}
			\frac{(S_1,\sigma)\Rightarrow\sigma'' (S,\sigma'')\Rightarrow\sigma'}{(S_1;S,\sigma)\Rightarrow\sigma'} 
		\end{equation*}
		
		Since $S_1\approx_{BS} S_2$, it is also true that $(S_2,\sigma)\Rightarrow\sigma''$, and therefore:
		
		\begin{equation*}
			\frac{(S_2,\sigma)\Rightarrow\sigma'' (S,\sigma'')\Rightarrow\sigma'}{(S_2;S,\sigma)\Rightarrow\sigma'}
		\end{equation*}
		
		By the definition of equivalence, both $(S_1,\sigma)$ and $(S_2,\sigma)$ result in the same state $\sigma''$, and therefore both $(S_1;S,\sigma)$ and $(S_2;S,\sigma)$ produce the same final state $\sigma'$.
				
	\item We can define the small-step semantics of \texttt{assert $b$ before $S$} using the following rule:
		\begin{equation*}
			\frac{}{(\texttt{assert }b\texttt{ before }S,\sigma)\rightarrow (S,\sigma)}eval(b,\sigma)=true
		\end{equation*}
		
		In other words, $S$ is executed if $b$ evaluates to $true$. Otherwise, no transition is produced.
		
		The difference between \texttt{assert $false$ before $S$} and \texttt{while $true$ do $skip$} is that computation finishes on the former case, despite the lack of a final state, while the latter case is an endless loop.
		
		The difference between \texttt{assert $false$ before $S$} and $skip$ can be illustrated by the sequential operation of two statements.	Consider a statement in the form of $(S_1;S_2)$ in which $S_1$ can be either \texttt{assert $false$ before $S$} or $skip$:
		
		\begin{equation*}
			\frac{}{(\texttt{assert }false\texttt{ before }S;S_2,\sigma)}	
		\end{equation*}
		
		\begin{equation*}
			\frac{}{(skip; S_2,\sigma)\rightarrow (S_2,\sigma)}	
		\end{equation*}

		It is possible to observe that, in the first case, the computation is aborted and $S_2$ is not executed. However, in the second case, $S_2$ is always executed with the same memory state as the one used by $skip$.
		
	\item Consider the statement:\\ \texttt{$x:= -1$; while $x\leq 0$ do ($x:=x-1$ or $x:= (-1)\times x$)}.
	For the purposes of brevity, let us call the second part of the statement as $S$.
	
	\begin{equation*}
		S = \texttt{while }x\leq 0\texttt{ do }(x:=x-1\texttt{ or }x:= (-1)\times x)
	\end{equation*}
	
	The \texttt{or} construct contained within $S$ leads to the existence of multiple possible final states for this statement. If the left-side path of the \texttt{or} is always chosen, no final state is produced and, therefore, there is no transition in big-step semantics.
	
	However, we can prove the transition for S using big-step semantics by assuming the right-side path of the \texttt{or} is taken and achieving a final state, for example:
	
	\begin{equation*}
		\frac{(x:=-1,\sigma)\Rightarrow\sigma[x\mapsto-1]\ \frac{(x:=(-1)\times x,\sigma[x\mapsto -1])\Rightarrow\sigma[x\mapsto1]\ (S;\sigma[x\mapsto 1])\Rightarrow\sigma[x\mapsto 1]}{(S,\sigma[x\mapsto -1])\Rightarrow \sigma[x\mapsto 1]}}{(x:=-1;S,\sigma)\Rightarrow\sigma[x\mapsto 1]}
	\end{equation*}
	
	In other words, given an initial state where $x$ is a negative integer $n$, following the right-side path of the \texttt{or} causes the following big-step transition: $(S,\sigma[x\mapsto n]) \Rightarrow \sigma[x\mapsto p]$, where $p = -n$.
	
	It is possible to observe that, by following the left-side path instead of the right-side path, the value of $x$ decreases and we find statement/state pairs such as $(S,\sigma[x\mapsto -2])$, $(S,\sigma[x\mapsto -3])$ and so forth. Since the left-side path can be taken an arbitrary number of times, for any negative integer $n$ there exists a transition in the form of $(S,\sigma[x\mapsto n])$. Then, by following the right-side path at any point, $x$ becomes positive and the computation ends.
	
	Therefore, this program as a whole can be described in big-step semantics as:
	
	\begin{equation*}
		\forall p\in \mathbb{N},\ (x:= -1; S,\sigma) \Rightarrow \sigma[x\mapsto p]
	\end{equation*}
		
	The development of the statement using small-step semantics is as follows:
	
		\begin{align*}
		(x:=-1; S,\sigma) \rightarrow \\
		(S,\sigma[x\mapsto-1]) \rightarrow \\
		(if\ x\leq 0\ then\ (x:=x-1\texttt{ or }x:= (-1)\times x); S\ else\ skip,\sigma[x\mapsto-1]) \rightarrow \\
		(x:=x-1\texttt{ or }x:= (-1)\times x; S,\sigma[x\mapsto-1]) \rightarrow
		\end{align*}

		From here, the following steps depend on the path taken by the \texttt{or}. Following the left-side path:
		
		\begin{align*}
		(x:=x-1; S,\sigma[x\mapsto-1]) \rightarrow \\
		(S,\sigma[x\mapsto-2]) \rightarrow \\
		(if\ x\leq 0\ then\ (x:=x-1\texttt{ or }x:= (-1)\times x); S\ else\ skip,\sigma[x\mapsto-2]) \rightarrow\\
		(x:=x-1\texttt{ or }x:= (-1)\times x; S,\sigma[x\mapsto-2]) \rightarrow\\
		(x:=x-1; S,\sigma[x\mapsto-2]) \rightarrow \\
		(S,\sigma[x\mapsto-3]) \rightarrow
		\end{align*}
		
		The development continues endlessly for all values of $x < 0$. Following the right-side path, we achieve a different conclusion:
		
		\begin{align*}
		(x:=(-1)\times x; S,\sigma[x\mapsto-1]) \rightarrow \\
		(S,\sigma[x\mapsto1]) \rightarrow \\
		(if\ x\leq 0\ then\ (x:=x-1\texttt{ or }x:= (-1)\times x); S\ else\ skip,\sigma[x\mapsto1]) \rightarrow \\
		(skip,\sigma[x\mapsto1]) \rightarrow \\
		\sigma[x\mapsto 1]
		\end{align*}
		
		While in big-step semantics this program always results in a state in which $x$ has the value of any natural number, in small-step semantics it is clear that the program might result in an endless loop.
		This illustrates the differences between the two approaches: the big-step semantics is useful to validate what the program does, but the small-step semantics allows us to see where execution could go wrong.
		
		Now, consider an extension to the language in the form of \texttt{random($x$)}, which produces a final state in which $x$ is a random natural number.
		Using big-step and small-step semantics, the transition provided by \texttt{random($x$)} is the same:
		
		\begin{equation*}
			\text{In big-step, }\forall p\in \mathbb{N},\ \frac{}{(\texttt{random}(x),\sigma)\Rightarrow\sigma[x\mapsto p]}
		\end{equation*}
		
		\begin{equation*}
			\text{In small-step, }\forall p\in \mathbb{N},\ \frac{}{(\texttt{random}(x),\sigma)\rightarrow\sigma[x\mapsto p]}
		\end{equation*}
		
		We have already seen a statement that produces an arbitrary natural number in big-step semantics. Therefore, when considering big-step semantics, \texttt{random($x$)} appears redundant.
		
		However, to show that \texttt{random($x$)} is redundant in small-step semantics, there needs to be a small-step transition that produces a random natural number without the risk of an endless loop. This is not possible using the \texttt{or} construct.
		
		Without the \texttt{random($x$)} function, the \texttt{or} construct is the only way to achieve nondeterminism in the While language. However, an \texttt{or} construct within a \texttt{while} loop effectively creates a binary tree, in which leaves are final states and nodes lead to more possible final states. Since there are infinite natural numbers, a statement that returns an arbitrary natural number must have infinite final states. Therefore, in small-step semantics, such a statement is represented by an infinite binary tree, which according to K\"onig's Lemma, always has an infinite path, making the statement not equivalent to \texttt{random($x$)}, which guarantees a final state is reached.
		
	\item The denotational semantics for \texttt{while $x\geq 0$ do $skip$} are as follows:
	\begin{equation*}
		g=\llbracket while\ x\geq 0\ do\ skip\rrbracket
	\end{equation*}
	\begin{equation*}
		F_{x\geq 0,skip}(g) = cond(eval(x\geq 0,-), g\circ \llbracket skip\rrbracket, id) = cond(eval(x\geq0,-),g,id)
	\end{equation*}
	\begin{equation*}
		F_{x\geq 0,skip}(g)(\sigma) = \begin{cases}\sigma\texttt{, if }\sigma(x)<0 \\ \bot\texttt{, if }\sigma(x) \geq 0\end{cases}
	\end{equation*}
	
	Consider a function $g_\bot$ such that $g_\bot(\sigma) = \bot$. Then:
	\begin{equation*}
		F_{x\geq 0,skip}(g_\bot) = cond(eval(x\geq0,-),g_\bot,id)
	\end{equation*}
	\begin{equation*}
		F_{x\geq 0,skip}(g_\bot)(\sigma) = \begin{cases}\sigma\texttt{, if }\sigma(x)<0 \\ g_\bot(\sigma) = \bot\texttt{, if }\sigma(x) \geq 0\end{cases}
	\end{equation*}
	
	\item 
	\begin{equation*}
		g=\llbracket repeat\ S\ until\ b\rrbracket=(FIX\ F_{!b,S})\circ \llbracket S\rrbracket
	\end{equation*}
	\begin{equation*}
		F_{!b,S}(g)=cond(eval(!b,-), g, id)
	\end{equation*}
	\begin{equation*}
		F_{!b,S}(g)(\sigma) = \begin{cases}\sigma\texttt{, if }eval(b)=true\\ \bot\texttt{, if }eval(b)=false\end{cases}
	\end{equation*}
		
	\end{enumerate}

\end{document}
